Descriptive, not injunctive, social norms caused increases in mask wearing during the COVID-19 pandemic

### 2023-01-25 manuscript.Rmd
changes before resubmission

author: 
  - name          : "Samantha L. Heiman"
    affiliation   : "*,1"
    #corresponding : yes
    #address       : "1101 E 10th St, Bloomington, IN 47405, United States"
    #email         : "slheiman@iu.edu"
  - name          : "Scott Claessens"
    affiliation   : "*,2"
  - name          : "Jessica D. Ayers"
    affiliation   : "3"
  - name          : "Diego Guevara Beltran"
    affiliation   : "4"
  - name          : "Andrew Van Horn"
    affiliation   : "5,6"
  - name          : "Edward R. Hirt"
    affiliation   : "1"
  - name          : "Athena Aktipis"
    affiliation   : "†,4"
  - name          : "Peter M. Todd"
    affiliation   : "†,1,7"

@@

abstract: |
 Human sociality is governed by two types of social norms: injunctive social norms, which prescribe what people _ought_ to do, and descriptive social norms, which reflect what people _actually_ do. These norms enable people to cooperate in the face of group-wide challenges. While previous experimental work has shown that people’s behavior is influenced by social norms, several open questions remain about the natural emergence of injunctive and descriptive social norms within populations and their causal influences on cooperative behavior over time. To understand how cooperative behavior emerges and is shaped by changing social norms in a non-experimental setting, we studied mask wearing during the COVID-19 pandemic. Mask wearing has individual benefits, but it is also a cooperative behavior that provides collective benefits of reduced disease transmission in the community. Leveraging two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = `r nrow(tar_read(d))`), we tracked people’s reported mask wearing and their perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends of norm perceptions and self-reported mask wearing suggested that norms and behavior were tightly coupled and both changed quickly in response to recommendations from public health authorities. In addition, a random-intercept cross-lagged panel model revealed that perceived descriptive norms, but not perceived injunctive norms, caused future within-person increases in individuals' mask wearing. These findings show that, during uncertain times, cooperative behavior is driven by what others are actually doing, rather than what others think ought to be done.

note              : "\\footnotesize \\raggedright \\mbox{* indicates shared first authorship, † indicates shared senior authorship} \\par Correspondence concerning this article should be addressed to Samantha L. Heiman, 1101 E 10th St, Bloomington, IN 47405, United States. E-mail: slheiman@iu.edu \\par This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation. \\par This working paper has not yet been peer-reviewed."
keywords          : "social norms; descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "5276 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

@@

# \normalfont Significance Statement

Social norms have been identified as important drivers of human cooperation, but the emergence of social norms within populations and their subsequent effects on cooperative behavior are not well understood. Here, we use mask wearing during the COVID-19 pandemic as a real-world setting in which to study the emergence of social norms. Over two years in the United States, we find that social norms and mask wearing are tightly coupled and change quickly in response to public health recommendations. Moreover, longitudinal modeling suggests that mask wearing is causally preceded by perceptions of descriptive norms (i.e. what people _are_ doing) but not injunctive norms (i.e. what people _ought_ to be doing).

@@

Social norms are a key aspect of human sociality [@Bicchieri2009; @Cialdini1991; @Fehr2018a; @VanKleef2019]. Broadly, social norms are defined as commonly known behavioral guidelines enforced by groups of people [@Legros2020]. Since Asch's early studies of normative conformity [@Asch1956], evidence has accrued that humans, unlike other species, are highly attuned to social norms [@Tomasello2014; @Riedl2012]. From a young age, children begin to adhere to and enforce group-wide social norms [@Jensen2014; @Rakoczy2013; @Schmidt2012]. Both children and adults rely on normative emotions, such as shame and guilt, to determine when they or others have violated social norms [@Vaish2011; @Schaumberg2022]. People also readily punish third-party norm violators [@Fehr2004]. Emerging from this uniquely normative psychology, social norms specify the rules and practices that govern human social life, forming the "grammar of society" [@Bicchieri2006].

By coordinating the behavior of many individuals, social norms enable human groups to cooperate in the face of group-wide challenges and threats, such as inter-group conflict, resource scarcity, high population density, natural disasters, and infectious diseases [@Gelfand2011; @Henrich2015; @Roos2015]. In the Maasai people of East Africa, for example, social norms governing transfers of livestock enable individuals to collectively manage the risks associated with nomadic pastoralism (e.g., famine, droughts) [@Aktipis2011]. In countries with high population densities, social norms provide the rules and regularities that allow large groups of people to effectively align their behavior and avoid chaos and conflict [@Gelfand2011]. Social norms are thus hypothesized to have played a key role in the evolution of large-scale cooperation in humans [@Henrich2015; @Chudek2011].

Previous research has distinguished between two primary types of social norms: injunctive norms and descriptive norms [@Bicchieri2009; @Cialdini1991; @Deutsch1955]. Injunctive social norms indicate what others in the group tend to approve or disapprove of and often involve social sanctions if violated. By contrast, descriptive norms simply describe what most people are doing in a given situation. Though these two kinds of social norms tend to align, they can also be in conflict with one another. For example, there may be an injunctive norm that cleaning up litter at a picnic site is the right thing to do: one _ought_ to behave this way. However, if an individual observes that most people are leaving their litter behind at the site, the descriptive norm is to not clean up. It is thus possible for injunctive and descriptive norms to differentially affect behavior [@Schultz2007].

Despite decades of research on the causes and consequences of injunctive and descriptive norms [@Bicchieri2009; @Cialdini1991; @Cialdini2007; @Cialdini2021; @Schultz2007], open questions remain regarding the emergence and causal influence of social norms [@Fehr2018a; @Legros2020; @VanKleef2019]. First, how do injunctive and descriptive norms emerge over time within a population? Second, how do evolving injunctive and descriptive norms causally influence behavior over time?

Research in cultural evolution and behavioral economics has investigated how social norms emerge in a population over time. In the long term, cultural evolutionary models show that injunctive social norms can be vertically transmitted across generations by imitation or teaching, or horizontally diffused from neighboring human populations [@Boyd1985; @Henrich2015]. For example, cultural phylogenetic studies have revealed patterns of vertical cultural inheritance across societies for a variety of injunctive social norms, such as norms governing land ownership [@Kushnick2014] and post-marital residence [@Jordan2009]. However, less is known about how social norms arise endogenously within populations in the short term. Recent experimental work in behavioral economics suggests that social norms of public good provisioning develop in tandem with cooperative behavior through repeated interactions [@Titlestad2019] and require peer enforcement to become stable [@Fehr2018b]. But it remains unclear whether these findings generalize beyond the laboratory to real human populations.

With regards to normative influences on behavior, there is a wealth of cross-sectional evidence demonstrating the impact of social norms on behavior. For example, field experiments in the United Kingdom and United States have demonstrated the positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. Evidence also suggests that any potentially deleterious effects of descriptive social norms (e.g., choosing to litter at a picnic site that already contains visible signs of littering) can be counteracted by instead focusing individuals’ attention on opposing injunctive norms (e.g., seeing that the litter is swept into piles, showcasing that it should be cleaned up) [@Cialdini1991; @Schultz2007].

However, these cross-sectional studies have two main limitations. First, studies have not adequately controlled for other potential non-social influences on behavior, such as perceptions of the effectiveness of the behavior and personal beliefs that one should behave in a certain way. Previous research has labeled these non-social beliefs as "factual beliefs" and "personal normative beliefs" respectively and emphasized that they should be distinguished from injunctive and descriptive social norms [@Bicchieri2014]. For example, willingness to sustainably reuse towels might be driven by perceptions that towel reuse has a positive impact on the environment and/or personal beliefs that towel reuse is the right thing to do, rather than by social norms. Second, cross-sectional studies have tended to follow experimental designs in which perceptions of social norms are manipulated by the researchers at a single time point. But social norms are not static: they change dynamically over time through processes of deliberation and interaction [@Smith2015]. An alternative way to assess causality between social norms and cooperative behavior, while retaining ecological validity, is to follow these variables over time amidst a real, unfolding social dilemma. In response to novel social dilemmas, social norms and cooperative behaviors emerge as people discover the benefits and costs of different behaviors [@Ostrom1990], enabling researchers to take multiple snapshots over time and establish temporal patterns. However, despite the promise of this approach, previous research has not studied the longitudinal interplay between social norms and cooperative behavior.

To understand how novel injunctive and descriptive social norms emerge over time and shape cooperative behavior in a non-experimental setting, in the current work we focus on mask wearing during the COVID-19 pandemic. Before the pandemic, mask wearing was not a common behavior in the United States. In April 2020, one month after the World Health Organization declared COVID-19 a global pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) as a protective behavior that people should adopt to minimize the spread of the disease [@CDC2022]. Mask wearing has individual benefits, but the CDC also emphasized the collective benefits of the behavior in reducing the spread of the disease throughout the community [@CDC2020]. Indeed, mask wearing posed a social dilemma to many individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the wider community (e.g., "flattening the curve" to protect at-risk individuals). Thus, the development of mask wearing in the United States during the COVID-19 pandemic allows us to study the emergence of novel injunctive and descriptive social norms and their causal effects on cooperative behavior over a short timescale within a single population.

Recent research has found positive relationships between individuals' perceptions of social norms and their protective COVID-19 behaviors. In the United States, one cross-sectional study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another vignette study found that experimentally-manipulated descriptive norms increased personal mask wearing intentions [@Bokemper2021]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are telling, but since they are experimental, cross-sectional, or only minimally longitudinal, they are unable to distinguish between within-person and between-person change over time [@Hamaker2015], nor do they have the temporal granularity to capture fluctuating changes in norm strength and norm adherence across the pandemic. Furthermore, these studies do not distinguish between the effects of social norms and the effects of other non-social beliefs that are predicted to shape behavior, such as factual beliefs and personal normative beliefs [@Bicchieri2014]. Several of the studies also do not control for political ideology, which is important to account for since mask wearing was highly politicized in the United States during the COVID-19 pandemic [@BaxterKing2022].

Here, we use two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. Between September 2020 and October 2022, we asked participants to report their frequency of mask wearing during in-person interactions, as well as their perceptions of descriptive and injunctive mask wearing norms. Guided by a causal model of social norms and mask wearing (Supplementary Figure \@ref(fig:plotDAG)), we also asked participants about their factual beliefs (i.e., whether they believe mask wearing is effective), personal normative beliefs (i.e., whether they believed mask wearing is the right thing to do), and political ideology, and controlled for these factors. We used our longitudinal data to answer our two main research questions in a specific real-world context. First, how do descriptive and injunctive mask wearing norms emerge and develop over time in the United States population during the pandemic? Second, how do descriptive and injunctive mask wearing norms causally influence mask wearing over time?

Second, fluctuations in mask wearing and perceived social norms are in line with recommendations broadcasted by the CDC, the main national public health agency of the United States. We do not have data for the very start of the pandemic in early 2020, but the high levels of mask wearing and strong perceived social norms at the start of our observation window likely emerged after the initial mask wearing recommendation from the CDC in April 2020. Perceived social norms and mask wearing subsequently declined after the CDC rescinded their mask wearing recommendation following widespread vaccine availability in March 2021, and then increased again after the CDC updated their guidelines for indoor mask use in high-risk areas in August 2021. Finally, perceived social norms and mask wearing declined again after the CDC eased mask wearing guidelines in March 2022. These trends were confirmed by a series of multilevel regression models with change points aligning with changes in CDC mask wearing recommendations (Supplementary Figure \@ref(fig:plotCDCSens); Supplementary Table \@ref(tab:changePointsTable)).

Our finding that social norms and mask wearing are tightly coupled over time provides real-world support for experimental evidence that social norms and cooperative behavior develop synergistically within groups via processes of social interaction [@Titlestad2019]. Moreover, the fact that these changes closely tracked the release of guidelines by the CDC supports the idea that institutions are part of the process by which culture and one’s own behaviors are mutually constructed [@Markus2010]. Indeed, empirical work in cultural evolution suggests that formal institutions are critical for the emergence and rapid adoption of novel social norms [@Amato2018]. While new norms can and do emerge spontaneously in populations, the process is slow compared to institution-driven norm change, which, as our descriptive trends have shown, can unfold over measurement intervals as short as four to six weeks.

In our longitudinal analysis, we found that descriptive norms, not injunctive norms, predicted future within-person increases in mask wearing across multiple time points. These cross-lagged effects were independent of the effects of non-social beliefs and political orientation. In line with this finding, similar descriptive norms have also been shown to predict future increases in physical distancing and prosocial behaviors (e.g., neighborhood help, charitable donations) during the COVID-19 pandemic, though this previous work did not adequately disentangle between-person and within-person effects as we have done here [@Rudert2021]. Similarly, experimental work has revealed that people in the United States are more likely to report intentions to wear a mask if they are told that others are wearing masks (i.e., a descriptive norm) [@Bokemper2021].

Why have descriptive norms had these positive effects on protective COVID-19 behaviors like mask wearing? Descriptive norms are particularly useful for coordinating behavior during fast changing, threatening situations with a high degree of uncertainty, such as the COVID-19 pandemic [@Gelfand2015]. During times of uncertainty, people look to others to get information quickly about how they should behave, and attempt to alleviate uncertainty-related stress by identifying with their group and its social norms [@Hogg2008; @Wellen1998]. Moreover, humans tend to be conditional cooperators who adapt their levels of cooperation depending on the degree of cooperative behavior in the population [@Chaudhuri2011]. Descriptive mask wearing norms provide evidence that others are cooperating, increasing the likelihood that individuals will themselves contribute to the public good by wearing masks.

Our longitudinal model revealed that, at the between-person level, people who perceived stronger injunctive norms on average also reported more frequent mask wearing on average. This correlation is in line with cross-sectional evidence showing that perceived injunctive norms are positively correlated with intentions to stay indoors during the pandemic among older adults [@Macy2021]. However, at the within-person level, we found that perceived injunctive norms did not directly predict future within-person increases in mask wearing, suggesting that injunctive norms and mask wearing are not directly causally related. This finding is at odds with previous evidence that experimentally-induced injunctive social norms can cause increases in cooperative behavior [@Schultz2007].

One possible explanation for these conflicting findings is that, unlike previous work, our model systematically controlled for non-social beliefs, such as factual beliefs and personal normative beliefs, which could potentially have driven previous experimental results. Another possible explanation is that, due to the increased opportunities to observe mask wearing in public, descriptive norms of mask wearing were made more salient than injunctive norms during the pandemic. According to the focus theory of normative conduct [@Cialdini1991], this difference in norm salience would produce behavior in line with descriptive norms and potentially suppress the effects of injunctive norms. By contrast, for more private behaviors like remaining indoors, it would have been less possible to observe other people’s behaviors, increasing the relative salience of injunctive norms. To test this idea, future research should expand our longitudinal cross-lagged approach to protective behaviors beyond mask wearing, including both public behaviors (e.g., physical distancing) and private behaviors (e.g., hand washing and home isolation).

While perceived injunctive norms did not directly predict future mask wearing, on one occasion they did indirectly predict future mask wearing _through_ perceived descriptive norms. In other words, perceptions of what others think people ought to be doing predicted future perceptions of what people were actually doing, which then predicted future mask wearing. This suggests that people may use injunctive norm information as a signal that they should check what others are doing, which could then aid in interpreting or perceiving other people's behavior as a descriptive norm. Future studies could manipulate injunctive norm presence and determine the effect of these norms on how people attend to and interpret behavioral information.

We are limited in generalizing these findings due to the constraints of our sample and the variables considered. While our sample began as representative of the United States, there was significant attrition over the course of the study (Supplementary Figure \@ref(fig:plotAttrition)). This attrition did not leave us with enough data to test the robustness of our results within different identity groups, such as different genders, ethnicities, or political ideologies. Since injunctive norm strength varies based on which group is seen as the source of the norm [@Neighbors2008], it would be interesting to learn whether different groups have different patterns of norm emergence over time. In particular, future analyses with larger samples should consider political ideology as a group identity, due to the political polarization of COVID-19 protective behaviors [@BaxterKing2022]. Our sample was predominantly White, and so future larger samples should be intentionally more diverse, answering calls to avoid generalizing White samples as representative of human behavior at large [@Remedios2022]. Our results also might not generalize to all social norms, behaviors, and social dilemmas. Norms governing sustainability in response to climate change, for example, might take longer to emerge, since the threat of climate change is more remote than the COVID-19 pandemic. For more distant social dilemmas that do not cause immediate day-to-day uncertainty, descriptive social norms may not necessarily drive cooperative behavior.

For the case of mask wearing in the United States during the COVID-19 pandemic, we have shown that social norms developed rapidly in the population and tracked ongoing changes in both recommendations from authorities and current levels of cooperative behavior. Moreover, we found that descriptive norms, rather than injunctive norms, were the main driver for future mask wearing. Importantly, this key finding slices two ways. Not only does it imply that high local levels of mask wearing encouraged future personal mask use, but it also implies that _low_ local levels of mask wearing _discouraged_ future personal mask use. This echoes recent reports of people in the United States not wanting to be "singled out" by being the only one wearing a mask in their community [@Natanson2022]. Our work thus underscores the importance of consistent, visible community adherence for encouraging personal protective behaviors in response to global pandemics like COVID-19.

@@

All analyses were conducted in R v4.1.1 [@RCoreTeam]. Visualizations were generated using the _cowplot_ [@Wilke2020] and _ggplot2_ [@Wickham2016] packages. The manuscript was reproducibly generated using the _targets_ [@Landau2021] and _papaja_ [@Aust2022] packages.

@@

This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation. We would also like to thank the Language, Culture, and Cognition lab at the University of Auckland for providing feedback on a previous version of this manuscript.

@@

SLH, JDA, DGB, AV, ERH, AA, and PMT conceptualized the study. SLH, SC, JDA, DGB, and AV oversaw the data curation, investigation, and methodology of the study. SLH and SC wrote the first draft of the paper. SC conducted the formal analysis and created all visualizations. ERH, AA, and PMT provided funding and supervision for the study. All authors reviewed and edited the final draft of the paper.

@@

# Research Transparency and Reproducibility

All data and code to reproduce the statistical analyses in this manuscript can be found on GitHub: https://github.com/ScottClaessens/covidMaskWearing

>>>>

author: 
  - name          : "Samantha L. Heiman"
    affiliation   : "*,1"
    #corresponding : yes
    #address       : "1101 E 10th St, Bloomington, IN 47405, United States"
    #email         : "slheiman@iu.edu"
  - name          : "Scott Claessens"
    affiliation   : "*,2"
  - name          : "Jessica D. Ayers"
    affiliation   : "3"
  - name          : "Diego Guevara Beltrán"
    affiliation   : "4"
  - name          : "Andrew Van Horn"
    affiliation   : "5,6"
  - name          : "Edward R. Hirt"
    affiliation   : "1"
  - name          : "Athena Aktipis"
    affiliation   : "†,4"
  - name          : "Peter M. Todd"
    affiliation   : "†,1,7"

@@

abstract: |
 Human sociality is governed by two types of social norms: injunctive norms, which prescribe what people _ought_ to do, and descriptive norms, which reflect what people _actually_ do. The process by which these norms emerge and their causal influences on cooperative behavior over time are not well understood. Here, we study these questions through social norms influencing mask wearing during the COVID-19 pandemic. Leveraging two years of data from the United States (18 time points; _n_ = `r nrow(tar_read(d))`), we tracked mask wearing and perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends suggested that norms and behavior were tightly coupled, changing quickly in response to public health recommendations. In addition, longitudinal modelling revealed that descriptive norms, not injunctive norms, caused future increases in mask wearing. During uncertain times, cooperative behavior is driven by what others are actually doing, rather than what others think ought to be done.

note              : "\\footnotesize \\raggedright \\mbox{* indicates shared first authorship, † indicates shared senior authorship} \\par Correspondence concerning this article should be addressed to Samantha L. Heiman, 1101 E 10th St, Bloomington, IN 47405, United States. E-mail: slheiman@iu.edu \\par This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation. \\par This working paper has not yet been peer-reviewed."
keywords          : "social norms; descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "4521 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
appendix          : "appendix.Rmd"
csl               : apa.csl
---

@@

# \normalfont Statement of Relevance

Social norms have been identified as important drivers of human cooperation, but the emergence of social norms within populations and their subsequent effects on cooperative behavior are not well understood. Here, we use mask wearing during the COVID-19 pandemic as a real-world setting in which to study the emergence of social norms. Over two years in the United States, we find that social norms and mask wearing are tightly coupled and change quickly in response to public health recommendations. Moreover, longitudinal modeling suggests that mask wearing is causally preceded by perceptions of descriptive norms (i.e. what people _are_ doing) but not injunctive norms (i.e. what people _ought_ to be doing). These findings show that visible community adherence is an important driver of mask wearing, suggesting potential avenues for behavior change in future global pandemics.

@@

Social norms are a key aspect of human sociality [@Bicchieri2009; @Cialdini1991; @Fehr2018a]. Broadly, social norms are defined as commonly known behavioral guidelines enforced by groups of people [@Legros2020]. By coordinating the behavior of many individuals, social norms enable human groups to cooperate in the face of group-wide challenges and threats, such as resource scarcity, natural disasters, and infectious diseases [@Roos2015]. Social norms are thus hypothesized to have played a key role in the evolution of large-scale cooperation in humans [@Henrich2015].

Previous research has distinguished between two types of social norms: injunctive norms and descriptive norms [@Bicchieri2009; @Cialdini1991; @Deutsch1955]. Injunctive norms indicate what others tend to approve or disapprove of and often involve social sanctions if violated. By contrast, descriptive norms simply describe what most people are doing in a given situation, but carry no prescriptive information _per se_. According to the focus theory of normative conduct [@Cialdini1991], these two kinds of social norms often align, but they can also be in conflict with one another and differentially affect behavior depending on which norm is more salient. For example, there may be an injunctive norm that cleaning up litter at a picnic site is the right thing to do: one _ought_ to behave this way. However, if an individual observes that most people are leaving their litter behind at the site, they may instead follow the descriptive norm and litter themselves.

Despite decades of research on injunctive and descriptive norms [@Cialdini1991; @Cialdini2021; @Schultz2007], open questions remain regarding the emergence and causal influence of social norms [@Legros2020; @VanKleef2019]. First, how do injunctive and descriptive norms emerge over time within a population? Second, how do evolving injunctive and descriptive norms causally influence behavior over time?

Research has investigated how social norms emerge in a population over time. In the long term, cultural evolutionary models show that injunctive social norms can be vertically transmitted across generations by imitation or teaching, or horizontally diffused from neighboring populations [@Henrich2015]. However, less is known about how social norms arise endogenously within populations in the short term. Recent work in behavioral economics suggests that social norms of public good provisioning develop in tandem with cooperative behavior through repeated interactions [@Titlestad2019]. But it remains unclear whether these findings generalize beyond the laboratory to real human populations.

With regards to normative influences on behavior, studies have demonstrated positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. However, these studies have two main limitations. First, studies have not accounted for other potential non-social explanations for behavior, such as perceptions of the effectiveness of the behavior and personal beliefs that one should behave in a certain way. These non-social beliefs, labeled "factual beliefs" and "personal normative beliefs" [@Bicchieri2014], often correlate with descriptive and injunctive norms, but they are fundamentally different because they can cause behavior separately from social expectations about what others do or think should be done. For example, willingness to recycle might be driven by perceptions that recycling has a positive impact on the environment and/or personal beliefs that recycling is the right thing to do, even if social norms actively discourage recycling (e.g., recycling is not a common behavior). It is thus important to control for non-social beliefs in studies of social norms. Second, studies have tended to follow cross-sectional experimental designs. However, social norms are not static: they change dynamically over time through processes of deliberation and interaction [@Smith2015]. An alternative but underutilized method of assessing causality between social norms and cooperative behavior, while retaining ecological validity, is to follow these variables over time amidst a real, unfolding social dilemma.

To understand how social norms emerge over time and shape cooperative behavior in a non-experimental setting, we focus on mask wearing in the United States during the COVID-19 pandemic. In April 2020, one month after the World Health Organization declared COVID-19 a global pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) to minimize the spread of the disease [@CDC2022]. Mask wearing has individual benefits, but the CDC also emphasized the collective benefits in reducing disease spread [@CDC2020]. Indeed, mask wearing posed a social dilemma to many individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the community (e.g., "flattening the curve" to protect at-risk individuals). Thus, the development of mask wearing during the COVID-19 pandemic enables us to study the emergence of social norms and their causal effects on cooperative behavior over a short timescale within a single population.

Recent research has found positive relationships between social norms and protective COVID-19 behaviors. In the United States, one study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another study found that experimentally-induced descriptive norms increased mask wearing intentions [@Bokemper2021]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are telling, but since they are cross-sectional or only minimally longitudinal, they do not have the temporal granularity to capture fluctuating changes in norm strength and adherence across the pandemic. Furthermore, several of the studies do not control for political ideology, which is important to account for since COVID-19 was highly politicized in the United States [@BaxterKing2022].

Here, we use two years of data from a representative sample of adults in the United States (18 time points; _n_ = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. Participants reported their frequency of mask wearing during in-person interactions, as well as their perceptions of descriptive and injunctive mask wearing norms. We also asked participants about their non-social mask wearing beliefs and political ideology, and controlled for these factors. We used these longitudinal data to answer two main research questions in a specific real-world context. First, how do descriptive and injunctive mask wearing norms emerge over time? Second, how do descriptive and injunctive mask wearing norms causally influence mask wearing?

@@

All analyses were conducted in R v4.1.1 [@RCoreTeam]. Visualizations were generated using the _cowplot_ [@Wilke2020] and _ggplot2_ [@Wickham2016] packages. The manuscript was reproducibly generated using the _targets_ [@Landau2021] and _papaja_ [@Aust2022] packages. All code and data are publicly available on GitHub [@Heiman2023].

Second, fluctuations in mask wearing and perceived social norms are in line with recommendations broadcasted by the CDC, an important institution governing public health in the United States. We do not have data for the very start of the pandemic in early 2020, but the high levels of mask wearing and strong perceived social norms at the start of our observation window likely emerged after the initial mask wearing recommendation from the CDC in April 2020. Perceived social norms and mask wearing subsequently declined after the CDC rescinded their mask wearing recommendation following widespread vaccine availability in March 2021, and then increased again after the CDC updated their guidelines for indoor mask use in high-risk areas in August 2021. Finally, perceived social norms and mask wearing declined again after the CDC eased mask wearing guidelines in March 2022. These trends were confirmed by a series of multilevel regression models with change points aligning with changes in CDC mask wearing recommendations (Supplementary Figure \@ref(fig:plotCDCSens); Supplementary Table \@ref(tab:changePointsTable)).

Our finding that social norms and mask wearing are tightly coupled over time provides real-world support for experimental evidence that social norms and cooperative behavior develop synergistically within groups via processes of social interaction [@Titlestad2019]. The fact that these changes closely tracked the release of guidelines by the CDC supports the idea that institutions are part of the process by which culture and one’s own behaviors are mutually constructed [@Markus2010]. Indeed, empirical work in cultural evolution suggests that formal institutions are critical for the emergence and rapid adoption of novel social norms [@Amato2018]. While new norms can and do emerge spontaneously in populations, the process is slow compared to institution-driven norm change, which, as our trends have shown, can unfold over measurement intervals as short as four to six weeks.

We found that descriptive norms, not injunctive norms, predicted future within-person increases in mask wearing, independent of the effects of non-social beliefs and political orientation. This finding is in line with previous evidence showing that perceptions of descriptive norms were positively related to other protective COVID-19 behaviors [@Bokemper2021; @Rudert2021]. There are several explanations for why descriptive norms have had these positive effects on protective COVID-19 behaviors like mask wearing. First, people may have followed descriptive norms to quickly coordinate their behavior with others during the pandemic. Descriptive norms are particularly useful for coordinating behavior during fast changing, threatening situations with a high degree of uncertainty, such as the COVID-19 pandemic [@Gelfand2015]. Second, people might have engaged in conditional cooperation, adapting their cooperation levels to the degree of cooperation in the population [@Chaudhuri2011]. Descriptive mask wearing norms provide evidence that others are cooperating, increasing the likelihood that individuals will themselves contribute to the public good by wearing masks. Third, the increased frequency of mask wearing in the population might have created a bandwagon effect [@SchmittBeck2015], encouraging conformist copying. Under this view, people wear masks not to coordinate or cooperate, but simply to fit in with the crowd. Future research will be required to determine the motivations underlying adherence to descriptive norms during uncertain times.

We found that perceived injunctive norms did not directly predict future within-person increases in mask wearing, suggesting that injunctive norms and mask wearing are not directly causally related. One possible explanation for this result is that, due to the increased opportunities to observe mask wearing in public, descriptive norms of mask wearing were more salient than injunctive norms during the pandemic. According to focus theory [@Cialdini1991], this difference in salience would produce behavior in line with descriptive norms and potentially suppress the effects of injunctive norms. By contrast, for more private behaviors like remaining indoors, it would have been less possible to observe other people’s behaviors, increasing the relative salience of injunctive norms. To test this idea, future research should expand our longitudinal approach to protective behaviors beyond mask wearing, including both public behaviors (e.g., physical distancing) and private behaviors (e.g., hand washing and home isolation).

We are limited in generalizing these findings due to the constraints of our sample and the variables considered. While our sample began as representative of the United States, there was significant attrition over the course of the study (Supplementary Figure \@ref(fig:plotAttrition)). This attrition did not leave us with enough data to test the robustness of our results within different identity groups, such as different genders, ethnicities, or political ideologies. Our results also might not generalize to all social norms, behaviors, and social dilemmas. Norms governing sustainability in response to climate change, for example, might take longer to emerge, since the threat of climate change is more remote than the COVID-19 pandemic. For more distant social dilemmas that do not cause immediate day-to-day uncertainty, descriptive social norms may not necessarily drive cooperative behavior.

For the case of mask wearing in the United States during the COVID-19 pandemic, we have shown that social norms developed rapidly in the population and tracked ongoing changes in both recommendations from authorities and current levels of cooperative behavior. Moreover, we found that descriptive norms, rather than injunctive norms, were the main driver for future mask wearing. Our work thus underscores the importance of consistent, visible community adherence for encouraging protective behaviors in response to global pandemics like COVID-19.

@@

[REDACTED]

@@

[REDACTED]

@@

# Open Practices Statement

All data and code to reproduce the statistical analyses in this manuscript are publicly available on GitHub: https://github.com/ScottClaessens/covidMaskWearing This study was not preregistered.
